Six ways to understand monads\\ \small{Experiences from the functional programming course at TU Delft}

### 2012-11-01 latex/paper.tex
Fill in text here and there

While functional programming has been taught for long in computer science
departments~\cite{Joost93}, curricula tend to emphasize functional programming
theory rather than practical applications. Special programming languages are
used to teach functional programming specific concepts, while little connection
is made to how those concepts can be transfered to non-functional languages.
The course at TU Delft attempted to teach functional programming theory in the
classroom and then expect students to apply those concepts in the (non-purely
functional) language of their choice. The course had a very strong teaching by
example focus: students were expected to participate in both in-classroom
exercises, homework assignments and implement a real world system as a final
project.

@@

The assignments were open ended, as there was no fixed set of targets for
each one of them. The deliverables for the assignments were a repository
with the source code and short report describing the solution that the
students came up with. The students were also required to have a working
demo by the end of the course time.

@@

@@

The student's response was overwhelming. Even though it was made clear that
the exercise would not contribute to the final grade, the will to apply the
data processing techniques taught during the lectures motivated
the students to work very hard. The fact that they were instructed to use
their favourite language enabled them to focus on decomposing the problem 
in a series of testable calculation steps rather than meddling with the
intricacies of a pure functional language. Solutions where provided in
Scala, {\sc c\#}, Scheme, Javascript and Haskell. 
Example renderings produced by student programs can be seen in 
Figure~\ref{fig:teapots}.

@@

@@

Coding sprints or ``hackathons'' have long been used by open source software
projects to speed up development and increase participation of community
members~\cite{Adams11}. In a typical sprint, project members set measurable
targets and work intensively to

@@

>>>>

While functional programming has been taught for long in computer science
departments~\cite{Joost93}, curricula tend to emphasize functional programming
theory rather than practical applications. Special programming languages are
used to teach functional programming specific concepts, while little connection
is made to how those concepts can be transfered to non-functional languages.
The course at TU Delft attempted to teach functional programming principles in the
classroom and then expect students to apply those concepts in the (non-purely
functional) language of their choice. The course had a very strong teaching by
example focus: students were expected to participate in both in-classroom
exercises, homework assignments and implement a real world system as a final
project.

@@

The assignments were open ended, as there was no fixed set of targets for
each one of them. The deliverables for the assignments were a repository
with the source code and short report describing the solution that the
students came up with. The students were also required to present a working
demo by the end of the course time.

@@

The focus of teaching was to present functional programming from a 
practical aspect.

The proverbial ``six ways to understand monads'' included presentations of the
formal theoretical concepts, visualizations of monads as blenders (based on the
monadic property that once a value is captured, it is then contained),
applications in Haskell for encapsulating state (through the \textsf{Maybe}
monad), applications to generic programming in {\sc c\#}, applications to {\sc
linq} and the reactive extensions framework, and, finally, a whiteboard exercise
where students where prompted to participate to the specification of Haskell's
\textsf{IO} monad from basic principles.

@@

The student's response was overwhelming. Even though it was made clear that the
exercise would not contribute to the final grade, the will to apply the data
processing techniques taught during the lectures motivated the students to work
very hard. The fact that they were instructed to use their favourite language
enabled them to focus on decomposing the problem in a series of testable
calculation steps rather than meddling with the intricacies of a pure functional
language.  Despite the advice, students were eager to test-drive the newly
acquired knowledge in a functional language; apart from {\sc c\#} and
Javascript, solutions were also provided in Scala, Scheme and Haskell. Example
renderings produced by student programs can be seen in Figure~\ref{fig:teapots}.

@@

According to many students, the Map/Reduce data analysis framework was what
drove them to attend the course. During the course, several examples
where presented

The result was that in their projects, students used implicit, through
the selected language's library collection methods, or explicit, by
formalizing data processing as a series of in memory Map/Reduce operations, forms
of Map/Reduce to process the data.

Out of personal interest, one student even implemented L\"ammel Map/Reduce
formulation in Scala, and through the language's extension mechanisms,
adapted it for use as an extension method by any Scala collection type,
including parallel collections.

@@

Coding sprints or ``hackathons'' have long been used by open source software
projects to speed up development and increase participation of community
members~\cite{Adams11}. In a typical sprint, project members set measurable
targets and work intensively in a tight deadline to deliver them. To ensure
the timely delivery of the student projects, we organized an one day
coding sprint, during the students where request to i) set an explicit target 
ii) write code for 7 hours with a short break iii) present a 5 minute 
demo of their work at the end of the day. All project teams where expected to
work in the same room, while communication among the teams was facilitated
by the lab demonstrator, to avoid unecessary distractions.

@@

As expected, the performance of students in project varied.
